\normalfont\spacedallcaps{The economic consequences of different attitudes of a policy maker: a combined epidemiological-econometric study}

### 2021-03-19 Manuscript/manuscript.tex
PUB
PUb

The recent pandemic due to the spread of the SARS-CoV-2 virus opened a highly debated issue about the best approach for the policy-maker to face the epidemic.  Unlike the past pandemics and the XX century (e.g., Spanish Flu,  Asiatic Flu, and Hong-Kong Flu), a massive amount of data are easily accessible for this pandemic.  Consequently, the modeling of the virus diffusion and the effect on socio-economic texture for different countries can be investigated with a more satisfactory resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally, the reproduction number, as described in Supporting Information).  Indeed the policy-maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme,  the spread of the virus is,  of course, stopped but,  on the other side,  the toll for such approach is that not only the economic activity (and so the income of people/firms) but also that the furniture of the primary goods is stopped.  On the other side,  if the PM takes no lock-down measures, the toll to be paid will be not only the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice, the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion to others (for this purpose, an excellent analysis was provided by Li et al. in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).
As a consequence, the lock-down efficacy,  within certain limits,  can be tuned.  In the literature, different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular, for the model in Ref. \cite{kabir2020evolutionary} it is assumed that the policy-maker knows the consequences of his/her choices ideally and that he/she can act without delay to impose the optimal choice; finally, it is assumed that the PM can impose a continuous factor for the lock-down (e.g., he/she can reduce the daily number of contacts between people by choosing each value between a definite range). In contrast, for different countries, such a factor seems to be much more discrete (e.g., the PM can reduce the daily contacts only by choosing definite values).  It can be argued that most of these drawbacks of this last formidable research are entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy-maker.  On this basis, we would propose here an alternative way to model the decision of the policy-maker that is based on a stochastic model instead of on a deterministic one.  Furthermore,  differently to the previous researches that focused basically on macroeconomic impact,  here we estimated and put in the model the impact of the different lock-downs at the microeconomic level: in particular, through the difference in difference,  we evaluated the effect of the different levels of lock-down on different sales sector as well as on subsidies (\textit{Cassa Integrazione Ordinaria}) in Italy.  Thus, the final output of the model will be the cumulative deaths and the economic damage for each selling sector.
Moreover, here we also considered that there is not only an economic cost for each death as done by \cite{kabir2020evolutionary}, but there is also average cost for each infected person (referring to Italian data) because a consistent part of them may be recovered or even should take the intensive therapy.  As we will show by varying with continuity the probabilistic parameter by which the PM imposes the lock-down,  a discontinuity in the SIRD model and the lock-down days was obtained.  Such behavior belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase, we will discuss the result of the simulation, and then we will compare them to get a general insight.

@@

where S is the number of people that are still susceptible, I the number of people that are infected, and R people that are recovered, while D are people that are dead.  N denotes the total population that for the timing of this paper it will be considered fixed\footnote{Othervise if longer horizontal timing is considered, it is necessary to consider a source term for the births and a well term for the natural deaths. For further details see \cite{vynnycky2010introduction}}. On the other side, $\beta$,$\gamma$, and $\mu$ are the parameters that shape the probability by which one individual in the model moves from a compartment to another: in particular, $\beta$ is the probability to be infected, $\gamma $ the probability to recover and $\mu$ the probability to die (basically the lethality defined as the probability to die conditionally to be ill).  Usually, epidemiologist is interested in the ratio:  

@@

The economic impact for each epidemiological scenario is shaped as follow: the first set of parameters,  as the economic value of death and of being infected by COVID-19, was taken directly from the reports/documentation of official sources; other parameters as the effect on selling for different areas and on unemployment benefit (Cassa Integrazione Guadagni) were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (300k EUR). This choice is based on the idea that,  if the PM misbehaves,  can be incriminated for manslaughter (with the consent of parliament that has to validate the incrimination) and then,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keep into account the age of the deaths: this in principle require an epidemiological model in which also the age of people is taken into account.  However, a partial differential equations system should be solved, making the calculation and the computational cost incredibly high.}.  Besides this impact,  there is also the cost associated with the medical care of each ill people. For this, we considered the average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another import health-care impact because the ill people for COVID-19 saturate the health system thus making it unreachable for other diseases.  This spillover translates into more death and more ill people for the baseline situation where there is not a pandemic: however,  by now,  this effect is difficult to quantify, and so we did not include it in the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the selling for the following ATECO-2007 \cite{Ateco2007} categories \footnote{In the rest part of the paper, these categories will be referred to as the part of the name labeled in blue. 

@@

The monthly sales data were retrieved from the National Institute of Statistics website \cite{istat}.  In particular, we considered the period starting from June 2018 up to June 2020 for the eight categories described in the previous section . The choice to not consider the months after June and, in particular, the last part of the year lies in the fact that in the latter lock-downs were imposed at the regional level and not to national level \footnote{With the only exception of the Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed on regional cluster making it more flexible: unfortunately the economic data used here, as far as the author knows,  were not available,  at all for a regional cluster.  Moreover, if regional clusters were considered, it was necessary to model an ensemble of SIRD models that communicate with a defined rate (that change also with respect to the lock-down restriction).  This makes the model much more complicated.  However, if all the necessary data for performing the analysis will become available, the author may consider, as an outlook, extending the present analysis to regional clusters}.  As discussed in Supporting Information,  these data were not de-seasoned. Thus, we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transform  \cite{huang1998empirical} for the decomposition of a time series data.  The decomposition results are reported in the Supporting Info E.  Concerning monthly subsidies paid we retrieved the data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}. Here the period considered also starts from June 2018 up to June 2020 for the four categories described in the previous section.  It is worth noting that we considered only the authorized subsidies and not the asked ones.   Furthermore,  differently for the sales, these data were not affected by seasonality noise,  and thus no de-seasoning was necessary.

@@

Now that we have the parameters for the economic impact given by the lock-downs,  we are ready to discuss the simulations that output the epidemiological and economic consequences of PM's attitude.  In Tab.  \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost}  the outcomes of the epidemiological model are given,  while the SIRD curves for the scenarios are provided in Fig.  \ref{SIRD_scenario}; on the other side the effect on selling and subsidies for both scenarios are given in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}.  To assure the \textit{ceteris pariubus} condition, we used the same set of random numbers for both scenarios.  First,  we see that where the PM is more reactive, we have different small epidemics waves. In contrast, if the PM is poorly reactive, only one intense wave is present,  indeed in the latter scenario, the laziness of PM to apply the lock-down in the first weeks produce the full infection peak of the standard SIRD model: when the PM acts is too late,  as shown in Fig.  \ref{SIRD_scenario}, since the pandemic has almost hit the large part of the population.  This can be pointed out in Fig. \ref{R_t_scenarios} by noting that that the reproduction factor remains at its maximum level for a large number of weeks after the beginning of the pandemic. This also explains why the lock-downs are so week in the second half of the scenario: the remaining part of the population is immune. Thus, no further action is required.  On the contrary, in the first scenario, the PM can use the lock-downs to transform the SIRD peak into small waves (Fig.  \ref{SIRD_scenario}).  As shown in Fig.  \ref{R_t_scenarios} this result is basically obtained by keeping the reproduction factor in the value of medium lock-down value.  These two theoretical models correspond,  in practice,  for the lazy PM to the one considered by the PMs that aimed to herd immunity. At the same time,  concerning the active-PM,  to the PMs that would not saturate the health system and aimed to reduce the death at the minimum.  It is interesting to point out that the change between the scenarios develops in a discontinuous way as shown in Fig.  \ref{Phase_transition}, as the reactivity of the PM is changed, we can consider this as a phase transition of a physical system (e.g., consider the gold-standard diagram of Ising model \cite{Ising,Ising_2}). This explains our choice to consider only two scenarios: indeed, we considered only a sample for each phase.  It is worth noting that a similar result,  within a different epidemiological model,  was obtained by Balcan and Vespignani in Ref.  \cite{Vespignani}.  The explanation of this similar behavior is that both models have a stochastic component,  that as pointed out by Balcan and Vespignani, gives the phase transition. While in their model, this was directly related to the contagion probability, this is indirectly made stochastically by PM's decision that modifies the $\beta$ parameter and so the transmission rate.  Moving to the socio-economic outcome of the model used, we see, from Tab. \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost},  that in the small wave scenario,  the attitude of the PM largely reduce the overall number of deaths and recovered (and so the cumulative cost for taking of patients) by consistent use of the confinement.  This has an immediate drawback on the economic data: as shown in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}, the losses for selling and the use of subsidies are widely large concerning the one-wave scenario.  In principle, one can be attempted to find the PM attitude that minimizes the overall cost (deaths,  infected,  selling, and subsidies): in the authors view, this scenario is not realistic since,  actually,  the Italian (but also many other European) criminal law does not allow this option: although there is an economic cost for life in terms of compensation,  the actual criminal law does not consider an amount of money compared to a money sum,  on contrary it gives a value ex-post, not ex-ante.

@@

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Manifattura.pdf}
         \caption{Subsidies paid for manufacture} 
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Commercio.pdf}}
         \caption{Subsidies paid for trade}
     \end{subfigure}
          \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Costruzioni.pdf}
         \caption{Construction}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_TOT.pdf}}
          \caption{Total}
     \end{subfigure}
\caption{Comparison between the ordinary subsidies (blue)   vs. the extraordinary ones (renovation red and purple solidarity) for a selected set of sectors and the overall total.  The dashed lines represent the different timing and intensity for the lock-down:  red for high, orange medium, and green for low.  }
\label{CIG_sector_plots}
\end{figure}

>>>>

The recent pandemic due to the spread of the SARS-CoV-2 virus opened a highly debated issue about the best approach for the policy-maker to face the epidemic.  Unlike the past pandemics of XX century (e.g., Spanish Flu,  Asiatic Flu, and Hong-Kong Flu), a massive amount of data are easily accessible for this pandemic.  Consequently, the modeling of the virus diffusion and the effect on socio-economic texture for different countries can be investigated with a more satisfactory resolution.  Among the different scientific challenges that can come up in this context,  an interesting one involves socio-economic effects of the attitude of the policy-maker (PM) to block the circulation of people (lock-down) in order to reduce the contagion rate (more formally, the reproduction number, as described in Supporting Information).  Indeed the policy-maker can adopt, at a first approximation,  a linear combination between these two extreme approaches: forcing all people to stay at home or to \textit{laissez-faire}.  In the first extreme,  the spread of the virus is,  of course, stopped but,  on the other side,  the toll for such approach is that not only the economic activity (and so the income of people/firms) but also that the furniture of the primary goods is stopped.  On the other side,  if the PM takes no lock-down measures, the toll to be paid will be not only the high number of deaths but also the economic damage produced by the very high number of deaths \cite{Correia2020,karlsson2014impact}.  In practice, the PM can adopt intermediate approaches that shut down activities that contribute much more to diffusion to others (for this purpose, an excellent analysis was provided by Li et al. in  \cite{li2021temporal} and by Brauner et al. in \cite{brauner2021inferring}).
As a consequence, the lock-down efficacy,  within certain limits,  can be tuned.  In the literature, different scholars \cite{kabir2020evolutionary,rowthorn2020cost} challenged the issue of finding the optimal lock-down policy for minimizing the economic impact as well as the deaths.  In particular, for the model in Ref. \cite{kabir2020evolutionary} it is assumed that the policy-maker perfectly knows the consequences of his/her choices and that he/she can act without delay to impose the optimal choice; finally, it is assumed that the PM can impose a continuous factor for the lock-down (e.g., he/she can reduce the daily number of contacts between people by choosing each value between a definite range). In contrast, for different countries, such a factor seems to be much more discrete (e.g., the PM can reduce the daily contacts only by choosing definite values).  It can be argued that most of these drawbacks of this last formidable research are entangled with the fact that a deterministic approach was considered for the activation of the lock-downs by the policy-maker.  On this basis, we would propose here an alternative way to model the decision of the policy-maker that is based on a stochastic model instead of on a deterministic one.  Furthermore,  differently to the previous researches that focused basically on macroeconomic impact,  here we estimated and put in the model the impact of the different lock-downs at the microeconomic level: in particular, through the difference in difference,  we evaluated the effect of the different levels of lock-down on different sales sector as well as on subsidies (\textit{Cassa Integrazione Ordinaria}) in Italy.  Thus, the final output of the model will be the cumulative deaths, the economic damage for each selling sector and the the increase of subsidies paid.
Moreover, here we also considered that there is not only an economic cost for each death as done by \cite{kabir2020evolutionary}, but there is also average cost for each infected person (referring to Italian data) because a consistent part of them may be recovered or even should take the intensive therapy.  As we will show by varying with continuity the probabilistic parameter by which the PM imposes the lock-down,  a discontinuity in the SIRD model and the lock-down days was obtained.  Such behaviour belongs to a class of phenomena that, in physical sciences, is called a phase transition.  For each phase, we will discuss the result of the simulation, and then we will compare them to get a general insight.

@@

where S is the number of people that are still susceptible, I the number of people that are infected, and R people that are recovered, while D are people that are dead.  N denotes the total population that for the timing of this paper it will be considered fixed\footnote{Othervise if longer horizontal timing is considered, it is necessary to consider a source term for the births and a well term for the natural deaths. For further details see \cite{vynnycky2010introduction}}. On the other side, $\beta$,$\gamma$, and $\mu$ are the parameters that shape the probability by which one individual in the model moves from a compartment to another: in particular, $\beta$ is the probability to be infected, $\gamma $ the probability to recover and $\mu$ the probability to die (basically the lethality defined as the probability to die given to be ill).  Usually, epidemiologist is interested in the ratio:  

@@

The economic impact for each epidemiological scenario is shaped as follow: the first set of parameters,  as the economic value of death and of being infected by COVID-19, was taken directly from the reports/documentation of official sources; other parameters as the effect on selling for different areas and on unemployment benefit (Cassa Integrazione Guadagni) were evaluated with empirical approaches from raw data.  Concerning the first set,  the number of deaths is multiplied by the maximum compensation value provided by the Court of Milan \cite{tribunaleMilano} for manslaughter (300k EUR). This choice is based on the idea that,  if the PM misbehaves,  can be incriminated for manslaughter (with the consent of parliament that has to validate the incrimination) and then,  if judged guilty,  charged by this amount for each death\footnote{In principle the judge also keep into account the age of the deaths: this in principle require an epidemiological model in which also the age of people is taken into account.  However, a partial differential equations system should be solved, making the calculation and the computational cost incredibly high.}.  Besides this impact,  there is also the cost associated with the medical care of each ill people. For this, we considered the average value calculated by National Anti-Corruption Authority (ANAC) \cite{Anac}: 28.180 kEUR \footnote{It is worth noting that,  in principle,  there is also another import health-care impact because the ill people for COVID-19 saturate the health system thus making it unreachable for other diseases.  This spillover translates into more death and more ill people for the baseline situation where there is not a pandemic: however,  by now,  this effect is difficult to quantify, and so we did not include it in the present model}.  Among the different sectors affected by the pandemic and the consequent lock-down,  we focused on the sales for the following ATECO-2007 \cite{Ateco2007} categories \footnote{In the rest part of the paper, these categories will be referred to as the part of the name labeled in blue. 

@@

The monthly sales data were retrieved from the National Institute of Statistics website \cite{istat}.  In particular, we considered the period starting from June 2018 up to June 2020 for the eight sales categories described in the previous section . The choice to not consider the months after June and, in particular, the last part of the year lies in the fact that in the latter lock-downs were imposed at the regional level and not to national level \footnote{With the only exception of the Christmas holiday}  \footnote{In principle one may ask why the present analysis was not performed on regional cluster making it more flexible: unfortunately the economic data used here, as far as the author knows,  were not available,  at all for a regional cluster.  Moreover, if regional clusters were considered, it was necessary to model an ensemble of SIRD models that communicate with a defined rate (that change also with respect to the lock-down restriction).  This makes the model much more complicated.  However, if all the necessary data for performing the analysis will become available, the author may consider, as an outlook, extending the present analysis to regional clusters}.  As discussed in Supporting Information,  these data were not de-seasoned. Thus, we performed a de-seasoning via the \textit{Forecast} R package \cite{forecast} that uses a Hilbert-Huang transform  \cite{huang1998empirical} for the decomposition of a time series data.  The decomposition results are reported in the Supporting Info E.  Concerning monthly subsidies paid we retrieved the data from the \textit{Osservatorio Cassa integrazione guadagni e fondi di solidarietà} on the \textit{Istituto nazionale della previdenza sociale} webpage \cite{inps}. Here the period considered also starts from June 2018 up to June 2020 for the four categories described in the previous section.  It is worth noting that we considered only the authorized (paid) subsidies and not the asked ones.   Furthermore,  differently for the sales, these data were not affected by seasonality noise,  and thus no de-seasoning was necessary.

@@

Now that we have the parameters for the economic impact given by the lock-downs,  we are ready to discuss the simulations that output the epidemiological and economic consequences of PM's attitude.  In Tab.  \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost}  the outcomes of the epidemiological model are given,  while the SIRD curves for the scenarios are provided in Fig.  \ref{SIRD_scenario}; on the other side the effect on selling and subsidies for both scenarios are given in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}.  To assure the \textit{ceteris pariubus} condition, we used the same set of random numbers for both scenarios.  First,  we see that where the PM is more reactive, we have different small epidemics waves. In contrast, if the PM is poorly reactive, only one intense wave is present,  indeed in the latter scenario, the laziness of PM to apply the lock-down in the first weeks produce the full infection peak of the standard SIRD model: when the PM acts is too late,  as shown in Fig.  \ref{SIRD_scenario}, since the pandemic has almost hit the large part of the population.  This can be pointed out in Fig. \ref{R_t_scenarios} by noting that that the reproduction factor remains at its maximum level for a large number of weeks after the beginning of the pandemic. This also explains why the lock-downs are so week in the second half of the scenario: the remaining part of the population is immune. Thus, no further action is required.  On the contrary, in the first scenario, the PM can use the lock-downs to transform the SIRD peak into small waves (Fig.  \ref{SIRD_scenario}).  As shown in Fig.  \ref{R_t_scenarios} this result is basically obtained by keeping the reproduction factor in the value of medium lock-down value.  These two theoretical models correspond,  in practice,  for the lazy PM to the one considered by the PMs that aimed to herd immunity. At the same time,  concerning the active-PM,  to the PMs that would not saturate the health system and aimed to reduce the death at the minimum.  It is interesting to point out that the change between the scenarios develops in a discontinuous way as shown in Fig.  \ref{Phase_transition}, as the reactivity of the PM is changed, we can consider this as a phase transition of a physical system (e.g., consider the gold-standard diagram of Ising model \cite{Ising,Ising_2}). This explains our choice to consider only two scenarios: indeed, we considered only a sample for each phase.  It is worth noting that a similar result,  within a different epidemiological model,  was obtained by Balcan and Vespignani in Ref.  \cite{Vespignani}.  The explanation of this similar behavior is that both models have a stochastic component,  that as pointed out by Balcan and Vespignani, gives the phase transition. While in their model, this was directly related to the contagion probability, this is indirectly made stochastically by PM's decision that modifies the $\beta$ parameter and so the transmission rate.  Moving to the socio-economic outcome of the model used, we see, from Tab. \ref{Table_Epidemiological_effect} and \ref{Table_Epidemiological_cost},  that in the small wave scenario,  the attitude of the PM largely reduce the overall number of deaths and recovered (and so the cumulative cost for taking of patients) by consistent use of the confinement.  This has an immediate drawback on the economic data: as shown in Fig. \ref{Selling_sector_effects} and \ref{CIG_sector_effects}, the losses for selling and the use of subsidies are widely large concerning the one-wave scenario.  In principle, one can be attempted to find the PM attitude that minimizes the overall cost (deaths,  infected,  selling, and subsidies): in the authors view, this scenario is not realistic since,  actually,  the Italian (but also many other European) criminal law does not allow this option (e.g. art. 452 Codice Penale): although there is an economic cost for life in terms of compensation,  the actual criminal law does not consider an amount of money compared to a money sum,  on contrary it gives a value ex-post, not ex-ante. 

@@

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Manifattura.pdf}
         \caption{Subsidies paid for manufacture} 
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Commercio.pdf}}
         \caption{Subsidies paid for trade}
     \end{subfigure}
          \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_Costruzioni.pdf}
         \caption{Construction}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{Figures/CIG_TOT.pdf}}
          \caption{Total}
     \end{subfigure}
\caption{Comparison between the ordinary subsidies (blue)   vs. the extraordinary ones (renovation red and purple solidarity) for a selected set of sectors and the overall total (including other sectors that were not analysed here).  The dashed lines represent the different timing and intensity for the lock-down:  red for high, orange medium, and green for low.  }
\label{CIG_sector_plots}
\end{figure}
